Value-Suppressing Uncertainty Maps

### 2017-09-17 paper/main.tex
Updated virus example, made more JH edits.

Next generation sequencing allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of relatively small numbers of base pairs. Using a dataset from a group of virologists interested in the population dynamics of viral mutation~\cite{o2012conditional}, we visualized the variability in viral genomes across 25 populations of the simian immunodeficiency virus (SIV). In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but also to not be distracted by false positives. The traditional bivariate map (\figref{fig:viral2d}) affords the analysis of only a few, large hot spots.

VSUPs (\figref{fig:viralVsum}) enable biologists to explore interesting patterns while deemphasizing noise. Smaller scale ``hot spots'' that are mapped to the same magenta color in the traditional map are visible as bright orange lines. Pink hot spots towards the end of the genome (around reference nucleotide $9,500$) could be considered potential regions of interest in the 2D map; in the VSUP these are encoded as highly uncertain and of ambiguous value. In fact, these regions have systematically low coverage; the lack of reliable data for these regions should discourage analysts from giving their apparent pattern equal consideration from the stronger signals earlier in the genome.

\textbf{A third example using a different visual form might be useful. For example, a node-link network with value + uncertainty associated with each node.}

@@

We limited our population to Turkers from the United States, with a prior task approval rate of at least 90\%. As the experimental tasks required multi-hue color perception, we presented participants with a set of Ishihara plates~\cite{hardy1945tests} as a pre-test, and excluded participants who either misidentified the values in the plates, or who self-reported as having a color vision deficiency (CVD) in the post-test. Based on piloting, we paid participants \$2 dollars for participation, for a target rate of \$8/hour. After completion of the main tasks, we solicited demographic information, including a risk aversion assay from Mandrik \& Bao~\cite{mandrik2005exploring}. We recruited a total of $48$ participants ($24$ for each experiment): $24$ female, $24$ male, ($M_{\text{age}}$= $34$, $SD_{\text{age}}$ = $9.1$). Participants from the first experiment were excluded from participating in the second.

%Rather than standard means, Cleveland \& McGill~\cite{cleveland1984graphical} use 25\% trimmed means to report effect sizes in graphical perception tasks. Trimmed means remove the first and last quartiles and then compute the mean from the remaining points. This filtering is important for graphical perception tasks (especially crowdsourced tasks) where error distributions may be long-tailed~\cite{heer2010crowdsourcing}. Trimmed means violate the sampling assumptions required for many null-hypothesis significance tests. Thus, while we report effect sizes in terms of trimmed means, we perform inferential statistical tests on the standard means. Asterisks in our charts denote significant differences, as reported by a post-hoc Tukey's test of Honest Significant Difference (HSD).

@@

Each of the 24 participants saw the data encoded as each of 8 types of bivariate encoding and legend, with 8 replications, for a total of 64 stimuli. \figref{fig:conditions} shows these factor levels.

@@

Our results from the prior study indicate that discrete, non-juxtaposed maps outperformed the other bivariate maps we selected, so we limited our study to only 4 types: square and wedge bivariate maps. These in turn were either VSUPs or traditional bivariate maps, for a 2 (square or wedge legend) x 2 (VSUP or standard) factorial design, with 6 replications, for a total of 24 stimuli for each of our 24 participants. \figref{fig:conditions} shows these factor levels. Prior to the main task, we included a short replication of the identification task from the prior experiment (with 12 stimuli) for training purposes. Poor performance on this training was used for exclusion purposes. Three people with low accuracy ($6\%$, $6\%$, and $25\%$ accuracy compared to a mean of $70\%$) were excluded.

@@

The results support our second hypothesis. An ANOVA for average danger value of guesses found a significant effect of quantization scheme ($F(1,61)=17$, $p<0.01$). A post-hoc t-test confirmed ($t=2.3$, $p=0.02$) that guesses made with VSUPs had significantly higher average danger ($M=0.32$, $SD=0.17$) (and so lower expected value) than traditional maps ($M=0.29$, $SD=0.18$).

>>>>

Next generation sequencing allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of relatively small numbers of base pairs. Using a dataset from a group of virologists interested in the population dynamics of viral mutation~\cite{o2012conditional}, we visualized the variability in viral genomes across 25 populations of the simian immunodeficiency virus (SIV). In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but also to not be distracted by false positives. The traditional bivariate map (\figref{fig:viral2d}) has many locations with apparent inter-population variance.

VSUPs (\figref{fig:viralVsum}) enable biologists to explore interesting patterns while deemphasizing noise. Orange hot spots towards the beginning (the second column) and end of the genome (around reference nucleotide $9,500$) could be considered potential regions of interest in the 2D map; in the VSUP these are encoded as highly uncertain and of ambiguous value. In fact, these regions have systematically low coverage; the lack of reliable data for these regions should discourage analysts from giving their apparent pattern equal consideration from the stronger signals elsewhere in the genome.

@@

We limited our population to Turkers from the United States, with a prior task approval rate of at least 90\%. As the experimental tasks required multi-hue color perception, we presented participants with a set of Ishihara plates~\cite{hardy1945tests} as a pre-test, and excluded participants who either misidentified the values in the plates, or who self-reported as having a color vision deficiency (CVD) in the post-test. Based on piloting, we paid participants \$2 dollars for participation, for a target rate of \$8/hour. After completion of the main tasks, we solicited demographic information, including a risk aversion assay from Mandrik \& Bao~\cite{mandrik2005exploring}. Participants from the first experiment were excluded from participating in the second.

@@

We recruited 24 participants for this task: $12$ female, $12$ male, ($M_{\text{age}}$= $31$, $SD_{\text{age}}$ = $7.8$). Each of the participants saw the data encoded as each of 8 types of bivariate encoding and legend, with 8 replications, for a total of 64 stimuli. \figref{fig:conditions} shows these factor levels.

@@

 We recruited 24 participants for this task: $12$ female, $12$ male, ($M_{\text{age}}$= $37$, $SD_{\text{age}}$ = $9.8$). Our results from the prior study indicate that discrete, non-juxtaposed maps outperformed the other bivariate maps we selected, so we limited our study to only 4 types: square and wedge bivariate maps. These in turn were either VSUPs or traditional bivariate maps, for a 2 (square or wedge legend) x 2 (VSUP or standard) factorial design, with 6 replications, for a total of 24 stimuli. Prior to the main task, we included a short replication of the identification task from the prior experiment (with 12 stimuli) for training and exclusion purposes. 3 people with unacceptably low accuracy on the training task ($6\%$, $6\%$, and $25\%$ accuracy compared to a mean of $70\%$) were excluded from analysis.

@@

The results support our second hypothesis. A similiar ANOVA for average danger value of guesses found a significant effect of quantization scheme ($F(1,59)=17$, $p<0.01$). A post-hoc t-test confirmed ($t=2.3$, $p=0.02$) that guesses made with VSUPs had significantly higher average danger ($M=0.32$, $SD=0.17$) (and so lower expected value) than traditional maps ($M=0.29$, $SD=0.18$).
